---
title: "Xbox eBay Auction Analysis: Semiparametric Regression"
date: "2026-05-24"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 5
)
```

# 1. Load Libraries & Data

```{r load}
library(tidyverse)
library(scales)
library(knitr)

# Load the dataset
df <- read.csv("Xbox_7-day_auctions.csv", stringsAsFactors = FALSE)

# Check the data
glimpse(df)
```

# 2. Basic Distributions

## 2.1 How Many Unique Auctions?

```{r unique_auctions}
n_auctions <- n_distinct(df$auctionid)
cat("Total unique auctions:", n_auctions, "\n")
cat("Total bid records   :", nrow(df), "\n")
```

## 2.2 Price & Opening Bid per Auction

```{r price_openbid}
# Each auction has a fixed price and openbid, so we collapse to one row per auction
auction_level <- df %>%
  group_by(auctionid) %>%
  summarise(
    openbid = first(openbid),
    price   = first(price),
    .groups = "drop"
  )

kable(auction_level, caption = "Closing Price & Opening Bid per Auction")
```

```{r price_openbid_plot}
# Price vs opening bid scatter
auction_level %>%
  ggplot(aes(x = openbid, y = price)) +
  geom_point(size = 3, colour = "#2c7bb6", alpha = 0.8) +
  geom_abline(linetype = "dashed", colour = "grey50") +  # 45° reference line
  scale_x_continuous(labels = dollar) +
  scale_y_continuous(labels = dollar) +
  labs(
    title    = "Closing Price vs Opening Bid by Auction",
    subtitle = "Dashed line = price equals opening bid",
    x        = "Opening Bid ($)",
    y        = "Closing Price ($)"
  ) +
  theme_minimal()
```

## 2.3 Bids per Auction

```{r bids_per_auction}
bids_per_auction <- df %>%
  group_by(auctionid) %>%
  summarise(n_bids = n(), .groups = "drop")

mean_bids <- mean(bids_per_auction$n_bids)

bids_per_auction %>%
  ggplot(aes(x = factor(auctionid), y = n_bids)) +
  geom_col(fill = "#2c7bb6", alpha = 0.8) +
  geom_hline(yintercept = mean_bids, color = "grey40", linetype = "solid", size = 1) +
  labs(
    title = "Number of Bids per Auction",
    x     = "Auction ID",
    y     = "Number of Bids"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

```

## 2.4 Bid Amounts

```{r bid_distribution}
summary(df$bid)

df %>%
  ggplot(aes(x = bid)) +
  geom_histogram(bins = 40, fill = "#2c7bb6", colour = "white", alpha = 0.85) +
  scale_x_continuous(labels = dollar) +
  labs(
    title = "Distribution of All Proxy Bids",
    x     = "Bid Amount ($)",
    y     = "Count"
  ) +
  theme_minimal()
```

## 2.5 Bid Timing — Days or Hours?

```{r bidtime_unit_check}
# bidtime is measured in DAYS from auction start (range 0–7)

cat("bidtime range (days):", range(df$bidtime), "\n")

# Create hours column for readability
df <- df %>%
  mutate(bidtime_hours = bidtime * 24)

cat("bidtime range (hours):", round(range(df$bidtime_hours), 1), "\n")
```

> We will keep the `bidtime` in **days** for modelling (natural scale of a 7-day auction),
> but plot in **hours** so patterns within a day are visible.

```{r bidtime_plot}
df %>%
  ggplot(aes(x = bidtime_hours)) +
  geom_histogram(bins = 60, fill = "#d7191c", colour = "white", alpha = 0.85) +
  scale_x_continuous(breaks = seq(0, 168, 24),
                     labels = paste0(seq(0, 7), "d")) +
  labs(
    title    = "When Are Bids Placed? (All Auctions)",
    subtitle = "x-axis labels show elapsed days; bars are ~2.8-hour bins",
    x        = "Time Since Auction Start (hours)",
    y        = "Number of Bids"
  ) +
  theme_minimal()
```

```{r bidtime_plot_by_auction}
df %>%
  ggplot(aes(x = bidtime_hours, fill = factor(auctionid))) +
  geom_histogram(bins = 60, colour = "white", alpha = 0.85, position = "stack") +
  scale_x_continuous(
    breaks = seq(0, 168, 24),
    labels = paste0(seq(0, 7), "d")
  ) +
  labs(
    title    = "When Are Bids Placed? (Stacked by Auction)",
    subtitle = "x-axis labels show elapsed days; bars are ~2.8-hour bins",
    x        = "Time Since Auction Start (hours)",
    y        = "Number of Bids"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

## 2.6 Bidder Rating Distribution

```{r bidderrate}
summary(df$bidderrate)

df %>%
  ggplot(aes(x = bidderrate)) +
  geom_histogram(bins = 40, fill = "#1a9641", colour = "white", alpha = 0.85) +
  scale_x_log10(labels = comma) +   # log scale because of extreme outliers
  labs(
    title    = "Distribution of Bidder Feedback Ratings (log scale)",
    subtitle = "Many new bidders (rating = 0); a few highly experienced buyers",
    x        = "Bidder Rating (log10)",
    y        = "Count"
  ) +
  theme_minimal()
```

## 2.7 Number of Unique Bidders per Auction

```{r unique_bidders}
bidders_per_auction <- df %>%
  group_by(auctionid) %>%
  summarise(n_bidders = n_distinct(bidder), .groups = "drop")

mean_bidders <- mean(bidders_per_auction$n_bidders)

bidders_per_auction %>%
  ggplot(aes(x = factor(auctionid), y = n_bidders)) +
  geom_col(fill = "#756bb1", alpha = 0.85) +
  geom_hline(
    yintercept = mean_bidders,
    color = "grey40",
    linetype = "solid",
    size = 0.6   # daha ince çizgi
  ) +
  labs(
    title = "Number of Unique Bidders per Auction",
    x     = "Auction ID",
    y     = "Unique Bidders"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))
```


# 3. Exploratory Data Analysis (EDA)

## 3.1 Unique Bidders, How Many Auction They Bid

```{r bidder_participation}
# How many unique bidders are there across all auctions?
n_unique_bidders <- n_distinct(df$bidder)
cat("Total unique bidders across all auctions:", n_unique_bidders, "\n")

# Count how many auctions each bidder participated in
bidder_participation <- df %>%
  group_by(bidder) %>%
  summarise(n_auctions = n_distinct(auctionid), .groups = "drop")

# Distribution: how many bidders participated in exactly 1, 2, 3... auctions
bidder_participation %>%
  count(n_auctions) %>%
  ggplot(aes(x = factor(n_auctions), y = n)) +
  geom_col(fill = "#2c7bb6", alpha = 0.85) +
  geom_text(aes(label = n), vjust = -0.5, size = 3.5) +
  labs(
    title    = "Bidder Participation Across Auctions",
    subtitle = paste("Total unique bidders:", n_unique_bidders),
    x        = "Number of Auctions Participated",
    y        = "Number of Bidders"
  ) +
  theme_minimal()
```



## 3.2 Bid Escalation — How Fast Do Bids Rise?

```{r bid_escalation}
# Show bid trajectory for a sample auction (most bids = most interesting)
top_auction <- bids_per_auction %>% slice_max(n_bids, n = 1) %>% pull(auctionid)

df %>%
  filter(auctionid == top_auction) %>%
  arrange(bidtime) %>%
  ggplot(aes(x = bidtime, y = bid, colour = bidder)) +
  geom_step(colour = "grey70") +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = dollar) +
  labs(
    title   = paste0("Bid Escalation — Auction ", top_auction),
    x       = "Time (days from auction start)",
    y       = "Proxy Bid ($)",
    colour  = "Bidder"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7))
```

## 3.3 Bidder Experience vs Bid Amount

```{r experience_vs_bid}
# Does a higher feedback rating lead to higher bids?
df %>%
  filter(bidderrate > 0) %>%   # exclude zero-rating bidders for log scale
  ggplot(aes(x = bidderrate, y = bid)) +
  geom_point(alpha = 0.3, colour = "#2c7bb6") +
  geom_smooth(method = "lm", colour = "#d7191c", se = TRUE) +
  scale_x_log10(labels = comma) +
  scale_y_continuous(labels = dollar) +
  labs(
    title    = "Bidder Rating vs Proxy Bid Amount",
    subtitle = "x-axis on log scale; OLS trend line with 95% CI",
    x        = "Bidder Feedback Rating (log10)",
    y        = "Proxy Bid ($)"
  ) +
  theme_minimal()
```

## 3.4 Price Markup Over Opening Bid

```{r markup}
auction_level <- auction_level %>%
  mutate(
    markup_abs = price - openbid,
    markup_pct_of_price = (price - openbid) / price * 100
  )

summary(auction_level %>% select(markup_abs, markup_pct_of_price))

auction_level %>%
  ggplot(aes(x = markup_pct_of_price)) +
  geom_histogram(bins = 20, fill = "#fc8d59", colour = "white", alpha = 0.85) +
  labs(
    title = "Closing Price Markup Over Opening Bid (%)",
    x     = "Markup (%)",
    y     = "Count"
  ) +
  theme_minimal()
```


## 3.8 Correlation Matrix (Numeric Variables)

```{r correlation}
library(corrplot)

num_vars <- df %>%
  select(bid, bidtime, bidderrate, openbid, price) %>%
  cor(use = "complete.obs")

corrplot(num_vars,
         method  = "color",
         type    = "upper",
         addCoef.col = "black",
         tl.col  = "black",
         tl.srt  = 45,
         title   = "Correlation Matrix",
         mar     = c(0, 0, 1, 0))
```

# 4. Summary Table

```{r summary_table}
# Per-auction summary combining everything
full_summary <- df %>%
  group_by(auctionid) %>%
  summarise(
    openbid       = first(openbid),
    close_price   = first(price),
    markup_pct_of_price    = round((first(price) - first(openbid)) / first(openbid) * 100, 1),
    n_bids        = n(),
    n_bidders     = n_distinct(bidder),
    median_bid    = median(bid),
    snipe_bids    = sum(bidtime > (7 - 1/24)),
    avg_bidrate   = round(mean(bidderrate), 1),
    .groups       = "drop"
  )

kable(full_summary,
      col.names = c("Auction ID", "Open ($)", "Close ($)", "Markup (%)",
                    "# Bids", "# Bidders", "Median Bid ($)",
                    "Snipe Bids", "Avg Bidder Rating"),
      caption = "Per-Auction Summary Statistics")
```
